<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{dpcR - package overview}
-->
`r library("knitr")`

# dpcR package

dpcR is an R package designed to assist in analysis of digital PCR experiments.

## Read data

dpcR does not have any inbuilt tools for reading specific data types. Instead, we advise to use other packages belonging do the pcRuniversum as [RDML](http://cran.r-project.org/web/packages/RDML/index.html) or dedicated packages such as ReadqPCR (available from bioconductor.org).

Before analysis data should be saved using one of the three classes - `adpcr` (end-point array digital PCR experiments), `ddpcr` (droplet digital PCR experiments) and `rtadpcr` (real-time array digital PCR experiments). To do so, use `create_dpcr` function.

```{r eval=TRUE}
library(dpcR)
#generate some data from 15x16 array. Let's presume, that we have results from two plates
sample_runs <- matrix(rpois(480, lambda = 1.5), ncol = 2)
#check its class - it's a typical R structure
class(sample_runs)
#save it to adpcr object
adpcr_experiments <- create_dpcr(sample_runs, n = 30L, type = "nm", adpcr = TRUE)
class(adpcr_experiments)
```

## Data visualisation and tests

Data can be easily visualized using `plot_panel` function.

```{r eval=TRUE}
#remember, you can plot only single panel at once 
plot_panel(extract_dpcr(adpcr_experiments, 1), nx_a = 15, ny_a = 16, main = "Experiment 1")
```

Randomness of spatial distribution on plate can be checked using `test_panel`.
```{r eval=TRUE}
#remember, you can plot only single panel at once 
test_panel(extract_dpcr(adpcr_experiments, 1), nx_a = 15, ny_a = 16)
```